---
title: "Indicator Species Repeated"
output:
  html_document:
    df_print: paged
  pdf_document: default
---
Based on the notebook I wrote for Rotem project.

If the data is not in long format start from here
```{r}
rm(list=ls()) # if need to start from scratch

library(tidyr)
library(reshape2)
 
  

data_wide <-as.data.frame(read.csv(file="IndicatorSpecies3.csv"))
head(data_wide)
data <- melt(data_wide, id='plots')  # 1st stage of becoming long
View(data)
class(data)

data_long <- data %>%  #It means - the following is done on the file data
separate(variable, into = c("year", "month"), sep = "\\.")  # 2nd stage, split the year and month

names(data_long)[4]<-"abundance" # rename the new col to the new name   dat$n
View(data_long)

## You should end up with a file for each species, with the exact same columns, just different number of lizards

#write.csv(data_long, file = "A.opheodurus.csv")
#write.csv(data_long, file = "S.doriae.csv")
write.csv(data_long, file = "S.sthenodactylus.csv")

```

!!!!! Now need to go to the saved file and add the standard columns, as in previous analyses

once the data is in long format- you can move on
The W at the end of the file stands for weighted, as this is the data file that has been weighted by number of surveyors - whenever 1 instead of 2 - results were multiplied by 2.

```{r}
dat=as.data.frame(read.csv(file="S.doriae.csv"))
dat <- dat[,1:6]

```



Now we can check normality assumptions. Start with visual inspection. Large deviations from the line, meaning not normal distribution. 

```{r}

# Visual inspection:
qqnorm(dat$Lizards)
qqline(dat$Lizards)


# To confirm our suspicions from visual inspection, we do a formal test

shapiro.test(dat$Lizards)

library(nortest)
lillie.test(dat$Lizards)

# If the p-value > 0.05, it is implying that the distribution of the data are not significantly different from normal distribution. In other words, we can assume the normality.
```

We see from the QQ plot and from the two tests that the data of lizards is very far from normality, thus we will analyze in non-parametric repeated measures ANOVA.
But because it cannot handle 3 within, we will remove the col. of year, thus keeping the months, but without lookng at differences between years.
we also need to average the two years, as it cannot handle 128 rows, when the number of combinations is 64. Used the apply in a two col. dataframe to calculate the averages below.


```{r}
dat16 <- subset(dat, Year==2016)
dat17<- subset(dat, Year==2017)
dat18 <- subset(dat, Year==2018)
JustLizards <- cbind(dat16$Lizards, dat17$Lizards, dat17$Lizards)
LizardsAVG <- apply(JustLizards,1,mean) 
datAVG <- cbind(dat16[,1:2],dat16[,4:5], LizardsAVG) # now we want to include the average as if it was the lizard number. We don't need the year, so we skip col. 3. We took dat16 as template, but each one would be the same.
View(dat)

```


```{r}
library(nparLD)

attach(datAVG)  # tells it which data set to use
length(datAVG$Plot)
ex.f1f2<-f1.ld.f2(y=LizardsAVG, time1=Month, group=Oil, time2=Treatment,   subject=Plot, time1.name="Month", group.name="Oil",
                   time2.name="Treatment",description=FALSE, time1.order=c(4,5,8,9),
                  group.order=c("new","old"), time2.order=c("T","C"))

ex.f1f2$ANOVA.test # automatically draws a plot of the main interaction, and a table of p values


```

Another version, from the article, but does the same, so can ignore.
```{r}
ex.f1f2np <- nparLD(LizardsAVG  ~ Oil  * Month* Treatment, data = datAVG,
 subject = "Plot", description = FALSE)
#plot(ex.f1f2np)
SummaryRanks <- summary(ex.f1f2np)
SummaryRanks
#print(ex.f1f2np)
                                                                                      
# write(SummaryRanks, file = "SummaryRanks.txt") # Write the list in a CSV file whose name is specified
# doesn't work, because it is not a table. So copy paste to a word or .txt file...
```


Now we will redo this, but showing only the years, and not the months (because we cannot analyze with three within with this package)
1. on 2016 - avearge the months
2. on 2017 - the same
3. on 2018 - the same
3. Combine the two again

```{r}
dat4 <- subset(dat16, Month==4)
dat5<- subset(dat16, Month==5)
dat8 <- subset(dat16, Month==8)
dat9<- subset(dat16, Month==9)
Lizards16 <- cbind(dat4$Lizards, dat5$Lizards,dat8$Lizards, dat9$Lizards) # average per year
Lizards16AVG <- apply(Lizards16,1,mean) # calculate the mean of the 4 months in each row

dat16AVG <- cbind(dat4[,1:3],dat4[,5], Lizards16AVG)  # combine for the 2016 dataset

# Now the same with 2017

dat4 <- subset(dat17, Month==4)
dat5<- subset(dat17, Month==5)
dat8 <- subset(dat17, Month==8)
dat9<- subset(dat17, Month==9)
Lizards17 <- cbind(dat4$Lizards, dat5$Lizards,dat8$Lizards, dat9$Lizards) # average per year
Lizards17AVG <- apply(Lizards17,1,mean)
dat17AVG <- cbind(dat4[,1:3],dat4[,5], Lizards17AVG)

# Now the same with 2018

dat4 <- subset(dat18, Month==4)
dat5<- subset(dat18, Month==5)
dat8 <- subset(dat18, Month==8)
dat9<- subset(dat18, Month==9)
Lizards18 <- cbind(dat4$Lizards, dat5$Lizards,dat8$Lizards, dat9$Lizards) # average per year
Lizards18AVG <- apply(Lizards18,1,mean)
dat18AVG <- cbind(dat4[,1:3],dat4[,5], Lizards18AVG)

# Now need to make sure the Lizard col. has the same name in both dataframes to be able to rbind them below
names(dat16AVG)[names(dat16AVG)=="Lizards16AVG"] <- "LizardsAVG"
names(dat17AVG)[names(dat17AVG)=="Lizards17AVG"] <- "LizardsAVG"
names(dat18AVG)[names(dat18AVG)=="Lizards18AVG"] <- "LizardsAVG"

datYearsAVG<- rbind(dat16AVG,dat17AVG,dat18AVG )

# now need to change the name of the 4th column, that somehow was changed during the cbind to "dat4[,5]", back to Treatment

names(datYearsAVG)[4] <- "Treatment"
View(datYearsAVG)
```


Now we can do the Repeated Measures to compare the means of the Years, without the Months

```{r}
rm(LizardsAVG, Year, Oil, Treatment, Plot)  # need to clear these, otherwise it seems to take some of the old ones, from the previous analysis, and then complains the lengths are different
attach(datYearsAVG)  # tells it which data set to use
ex.f1f2<-f1.ld.f2(y=LizardsAVG, time1=Year, group=Oil, time2=Treatment,   subject=Plot, time1.name="Year", group.name="Oil",
                   time2.name="Treatment",description=FALSE, time1.order=c(2016,2017,2018),
                  group.order=c("new","old"), time2.order=c("T","C"))

ex.f1f2$ANOVA.test # automatically draws a plot of the highest interaction (even if not significant, so ignore), and a table of p values
```

Finally we need to get the numbers of lizards in each type, in order to plot the results in a visible way. Could plot the ranks, which seems to create similar shape of plots, but # of lizards is more intuitive. Actually, need to call them # of observations...
So will do this by calculating ezANOVA and ask for the means.
Ignore the first part, because we know canot do the parametric ANOVA. But in chunks below they  calculate the means

```{r}
library(ez)

dat=as.data.frame(read.csv(file="S.doriae.csv"))  # NEED TO MAKE SURE TAKING THE SAME SPECIES AS ABOVE
dat <- dat[,1:6]

ezPrecis(dat)  # if you look here - Year and Month are numeric, and should be factors, otherwise assumptions are not tested. So below we convert them to factors.

dat[,"Year"] <- factor(dat[,"Year"])
dat[,"Month"] <- factor(dat[,"Month"] )
ezPrecis(dat) # check again that all is OK

Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year,Month,Treatment)
, between = Oil
)

print(Regular_Anova)



```
Can see that the Sphericity Corrections make almost everything not significant. But remember the assumptions of normality were seriously violated, that's why we did the non-parametric.


Calculate below means for the between factor - Oil (age) - remember it mixes treatment, so it is mostly habitat 
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, between = .(Oil)
)

MeansYear <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, between = .(Oil)
)

MeansYear
write.csv(MeansYear, file = "MeansOil.csv") # Write the dataframe in a CSV file whose name is specified
```

Calculate below means for the year 
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year)
)

MeansYear <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year)
)

MeansYear
write.csv(MeansYear, file = "MeansYear.csv") # Write the dataframe in a CSV file whose name is specified
```

Means for the months
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Month)
)

MeansMonths <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Month)
)
MeansMonths
write.csv(MeansMonths, file = "MeansMonths.csv") # Write the dataframe in a CSV file whose name is specified
```
for interaction oil*month
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Month)
, between = Oil
)

MeansOil.Treatment <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Month)
, between = Oil
)

MeansOil.Treatment
write.csv(MeansOil.Treatment, file = "MeansOil.Month.csv")
```

for interaction oil*Year
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year)
, between = Oil
)

MeansOil.Treatment <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year)
, between = Oil
)

MeansOil.Treatment
write.csv(MeansOil.Treatment, file = "MeansOil.Year.csv")
```


for interaction oil*treatment
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Treatment)
, between = Oil
)

MeansOil.Treatment <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Treatment)
, between = Oil
)

MeansOil.Treatment
write.csv(MeansOil.Treatment, file = "MeansOil.Treatment.csv")
```

for interactions year*treatment
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year,Treatment)
, between = NULL
)

MeansYear.Treatment <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year,Treatment)
, between = NULL
)
MeansYear.Treatment
write.csv(MeansYear.Treatment, file = "MeansYear.Treatment.csv")
```

Now three way interactions, like the plots created by the non-parametric
```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year,Treatment)
, between = Oil
)

MeansOil.Year.Treatment <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Year,Treatment)
, between = Oil
)
MeansOil.Year.Treatment
write.csv(MeansOil.Year.Treatment, file = "MeansOil.Year.Treatment.csv")


```
The other three way interactions

```{r}
Regular_Anova <- ezANOVA(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Month,Treatment)
, between = Oil
)

MeansOil.Month.Treatment <- ezStats(
data=dat
, dv=Lizards
, wid=Plot
, within = .(Month,Treatment)
, between = Oil
)
MeansOil.Month.Treatment
write.csv(MeansOil.Month.Treatment, file = "MeansOil.Month.Treatment.csv")


```








For comparison - the original data set in package

```{r}
data(ANT)
head(ANT)
ezPrecis(ANT)

#Run an ANOVA on the mean correct RT data.
rt_anova = ezANOVA(
data = ANT[ANT$error==0,]
, dv = rt
, wid = subnum
, within = .(cue,flank)
, between = group
)
#Show the ANOVA and assumption tests.
print(rt_anova)
```






